Degeneracy in a restricted Boltzmann machine

### 2016-02-03 writing/degeneracy/draft.Rmd
more writing

The mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$, is defined as
\begin{align*}
\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ t(\boldsymbol X) \right] \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
\end{align*}

@schweinberger2011instability introduces a related concept, instability, which  is characterized
by excessive sensitivity and near-degeneracy. He provides the following tool for detecting instability in the parameters of a discrete exponential family model, and thus near-degeneracy.

\begin{defn}
A sufficient statistic $t_k(\boldsymbol x_N)$ is \emph{stable} if there exist constants $C > 0$ and $N_C > 0$ such that
\begin{align}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N} [t_k(\boldsymbol x_N)] \le CN \text{ for all } N > N_C
\end{align}
and \emph{unstable} if for any $C > 0$, however large, there exists $N_C > 0$ such that
\begin{align}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N} [t_k(\boldsymbol x_N)] > CN \text{ for all } N > N_C
\end{align}
\end{defn}

\begin{lemma}
A $K$-parameter exponential family $\{P_{\theta} , \theta \in \Theta \}$ with natural parameters $\theta_1,\dots, \theta_K$ and $K - 1$ stable sufficient statistics, $t_1(\boldsymbol x_N),\dots, t_{K-1}(\boldsymbol x_N)$, as well as one unstable sufficient statistic, $t_K(\boldsymbol x_N)$, is degenerate with respect to $t_K(\boldsymbol x_N)$.
\end{lemma}

@@

expand.grid(theta_h1 = seq(-2, 2, length.out = 5),
            theta_v1 = seq(-2, 2, length.out = 5),
            theta_11 = seq(-2, 2, length.out = 11)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
                mu_v1 = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_h1 = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_11 = possibles[,"theta11"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
  mutate_each(funs(./gamma), starts_with("mu")) %>%
  select(-gamma) %>%
  gather(mu_parameter, value, mu_v1, mu_h1, mu_11) %>%
  mutate(theta_h1 = factor(theta_h1, labels = paste0("theta[h1]==", unique(theta_h1))), 
         theta_v1 = factor(theta_v1, labels = paste0("theta[v1]==", unique(theta_v1)))) %>%
  ggplot() +
  geom_line(aes(theta_11, value, colour = mu_parameter)) +
  facet_grid(theta_h1 ~ theta_v1, labeller = label_parsed) +
  xlab(expression(theta[11])) +
  scale_colour_discrete(expression(mu(theta)), labels = c(expression(theta[v1]), 
                                                          expression(theta[h1]), 
                                                          expression(theta[11])))

@@

```{r split_plots}
plot_dat_grid %>%
  group_by(r1, r2, n_param) %>%
  summarise(frac_degen = sum(near_hull)/n(), count = n()) %>% 
  ungroup() %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = frac_degen)) +
  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", bins = 8) +
  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", breaks = .05, size = 1.5) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(low = "yellow", high = "red") +
#   geom_point(aes(r1, r2, alpha = frac_degen), 
#              data = plot_dat_split %>%
#                group_by(r1, r2, n_param) %>%
#                summarise(frac_degen = sum(near_hull)/n(), count = n())) +
  facet_grid(~n_param) +
  theme(aspect.ratio = 1) +
  theme(legend.position = "bottom")

plot_dat_grid %>% 
  mutate(ss_interaction = sqrt(ss_interaction)/(H*V), ss_main = sqrt(ss_main)/(H + V)) %>%
  #filter(near_hull) %>%
  ggplot() +
  geom_density_2d(aes(ss_main, ss_interaction, colour = ..level..)) +
  scale_colour_gradient(low = "yellow", high = "red") +
  facet_grid(~n_param) +
  theme(aspect.ratio = 1) +
  theme(legend.position = "none")

>>>>

The mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$, for the RBM is defined as
\begin{align*}
\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ t(\boldsymbol X) \right] \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
\end{align*}

@schweinberger2011instability introduces a related concept, instability, which  is characterized
by excessive sensitivity and near-degeneracy. He provides the following tool for detecting instability in the a discrete exponential family model, and thus near-degeneracy.

\begin{defn}
A discrete exponential family distribution is unstable if for any $C > 0$, however large, there exists $N_C > 0$ such that 
\begin{align}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN \text{ for all } N > N_C \label{eq:unstable}
\end{align}
\end{defn}

\begin{lemma}
If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X_N \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N}$ is the subset of $\epsilon$-modes of the probability mass function.
\end{lemma}

In the context of the RBM, $N$ is the number of nodes in the model, $H + V$. So, the distribution is unstable if for any $C > 0$,

\begin{align*}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] &= \max\limits_{\boldsymbol x \in \mathcal{X}}\left[ \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j \right] \\
&= \max\limits_{h_j, v_i \in \{0,1\}}\left[ \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j \right] \\
&> C(H + V)
\end{align*}

There are $H + V + H*V$ terms in the sum that comprises $Q(\boldsymbol x)$, which leads to a natural split of the quantity into main effects, those terms based solely on $h_i$ and $v_j$ and interaction effects, those terms based on the product of $h_i v_j$. There are $H+V$ main effects terms and $H*V$ interaction effects terms. For $H$ and $V$ large enough, the number of interaction terms is growing much faster than $H + V$, meaning no matter the $C$, we are able to find values of $\boldsymbol theta$ that can make the left side of equation \ref{eq:unstable} large compared to $C(H+V)$. This points to the interaction terms as potentially the more dangerous parameters for causing near-degeneracy in the RBM.

```{r interaction size, include=FALSE}
expand.grid(H = 1:10, V = 1:10) %>%
  mutate(interaction_terms = H*V) %>%
  ggplot() +
  geom_contour(aes(H, V, z = interaction_terms/(H + V), colour = ..level..), bins = 20) +
  geom_contour(aes(H, V, z = interaction_terms/(H + V)), colour = "black", breaks = 1) +
  scale_colour_gradient("H*V/(H + V)", low = "yellow", high = "red")

@@

expand.grid(theta_h1 = seq(-2, 2, length.out = 5),
            theta_v1 = seq(-2, 2, length.out = 5),
            theta_11 = seq(-2, 2, length.out = 11)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
                mu_v1 = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_h1 = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_11 = possibles[,"theta11"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
  mutate_each(funs(./gamma), starts_with("mu")) %>%
  select(-gamma) %>%
  gather(mu_parameter, value, mu_v1, mu_h1, mu_11) %>%
  ungroup() %>%
  mutate(theta_h1 = factor(theta_h1, labels = paste0("theta[h1]==", unique(theta_h1))), 
         theta_v1 = factor(theta_v1, labels = paste0("theta[v1]==", unique(theta_v1)))) %>%
  ggplot() +
  geom_line(aes(theta_11, value, colour = mu_parameter)) +
  facet_grid(theta_h1 ~ theta_v1, labeller = label_parsed) +
  xlab(expression(theta[11])) +
  scale_colour_discrete(expression(mu(theta)), labels = c(expression(theta[v1]), 
                                                          expression(theta[h1]), 
                                                          expression(theta[11])))

@@

To explore the behavior of the RBM parameters $\boldsymbol \theta$ as it related to model degeneracy and near degeneracy, we example various models of small size. For $H, V \in \{1, \dots, 4\}$, we simulate `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitude, $||\boldsymbol \theta || = \sqrt{\sum_{i = 1, \dots, H + V + H*V} \theta_i^2}$, between `r min(plot_dat_grid$r1)` and `r max(plot_dat_grid$r1)`. For each set of parameters we then calculate the mean parameterization, as detailed in section \ref{degeneracy-in-random-graph-models} and can classify each model as near-degenerate or viable and look at the fraction of models that were near-degenerate for each combination of magnitude of $\boldsymbol \theta$ and model size.

In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to, and allow the two types of terms to have varying magnitudes. The results of this numerical study can be seen in figure \ref{fig:split_plots}.

```{r split_plots, fig.cap="\\label{fig:split_plots}A numerical experiment looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size."}
plot_dat_grid %>%
  group_by(r1, r2, H, V) %>%
  summarise(frac_degen = sum(near_hull)/n(), count = n()) %>% 
  ungroup() %>% 
  mutate(Hiddens = paste0("Hiddens: ", H), Visibles = paste0("Visibles: ", V)) %>%
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = frac_degen)) +
  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", bins = 8) +
  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", breaks = .05, size = 1.5) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient("Fraction Degenerate", low = "yellow", high = "red") +
#   geom_point(aes(r1, r2), alpha = .1, 
#              data = plot_dat_split %>%
#                mutate(ss_interaction = sqrt(ss_interaction), ss_main = sqrt(ss_main)) %>%
#                select(-r1, -r2) %>%
#                rename(r1 = ss_main, r2 = ss_interaction) %>%
#                group_by(r1, r2, n_param) %>%
#                summarise(frac_degen = sum(near_hull)/n(), count = n()) %>%
#                filter(frac_degen > 0)) +
  facet_grid(Visibles~Hiddens) +
  xlab(expression(group("||", theta[main], "||"))) +
  ylab(expression(group("||", theta[interaction], "||"))) +
  theme(aspect.ratio = 1)
  #theme(legend.position = "bottom")
```

```{r split_plots2, warning=FALSE}
plot_dat_grid %>% 
  mutate(dist = sqrt(r1^2 + r2^2)) %>%
  group_by(dist, H, V) %>%
  summarise(frac_degen = sum(near_hull)/n()) %>%
  mutate(Hiddens = as.character(H), Visibles = paste0("Visibles: ", V)) %>%
  ggplot() +
  geom_point(aes(dist, frac_degen, colour = Hiddens)) +
  geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
  facet_grid(~Visibles) +
  ylim(c(0,1)) +
  xlab(expression(paste(group("||", theta, "||"), ", Distance from the origin"))) +
  ylab("Fraction Degenerate")
```
